library(tidyverse)
library(ggsurvfit)
library(dplyr)
library(survival)
library(survminer)

setwd('../Survival-Analysis-DSTI-Project')
gym_data <- read_csv('datas/gym_churn_us.csv')

gym_data$censored <- ifelse(gym_data$Churn == 0, TRUE, FALSE)

surv_object <- Surv(time = gym_data$Lifetime, event = gym_data$Churn)

km_fit <- survfit(surv_object ~ 1,data=gym_data)
km_summary <- summary(km_fit)
print(paste("survival probabilities: ", min(km_summary$surv)))
print(paste("survival times: ", max(km_summary$time)))

```{r}
km_fit_transplant <- survfit(surv_object ~ Contract_period, data = gym_data)
#summary(km_fit_transplant) # commented out to reduce the pages
ggsurvplot(km_fit_transplant, conf.int = TRUE, ggtheme = theme_minimal(),
           title = "Survival Curve by Transplant Status with Censoring", 
           censor.shape = '|', censor.size=4)
```

```{r}
km_fit_transplant <- survfit(surv_object ~ Month_to_end_contract, data = gym_data)
#summary(km_fit_transplant) # commented out to reduce the pages
ggsurvplot(km_fit_transplant, conf.int = TRUE, ggtheme = theme_minimal(),
           title = "Survival Curve by Transplant Status with Censoring", 
           censor.shape = '|', censor.size=4)
```

```{r}
gym_data <- within(gym_data, {   
  Age.cat <- NA # need to initialize variable
  Age.cat[Age < 25] <- "18-25"
  Age.cat[Age >= 25 & Age < 30] <- "25-30"
  Age.cat[Age >= 30  & Age < 35] <- "30-35"
  Age.cat[Age >= 35] <- "35+"
   } )
```

```{r}
km_fit_transplant <- survfit(surv_object ~ Age.cat, data = gym_data)
#summary(km_fit_transplant) # commented out to reduce the pages
ggsurvplot(km_fit_transplant, conf.int = TRUE, ggtheme = theme_minimal(),
           title = "Survival Curve by Transplant Status with Censoring", 
           censor.shape = '|', censor.size=4)
```

```{r}
gym_data <- within(gym_data, {   
  Month_to_end_contract.cat <- NA # need to initialize variable
  Month_to_end_contract.cat[Month_to_end_contract >= 1 & Month_to_end_contract < 3] <- "1-2"
  Month_to_end_contract.cat[Month_to_end_contract >= 3  & Month_to_end_contract < 6] <- "3-5"
  Month_to_end_contract.cat[Month_to_end_contract >= 6] <- "> 6"
   } )
```

```{r}
km_fit_transplant <- survfit(surv_object ~ Month_to_end_contract.cat, data = gym_data)
#summary(km_fit_transplant) # commented out to reduce the pages
ggsurvplot(km_fit_transplant, conf.int = TRUE, ggtheme = theme_minimal(),
           title = "Survival Curve by Transplant Status with Censoring", 
           censor.shape = '|', censor.size=4)
```

```{r}
gym_data <- within(gym_data, {   
  Avg_additional_charges_total.cat <- NA # need to initialize variable
  Avg_additional_charges_total.cat[Avg_additional_charges_total >= 0 & Avg_additional_charges_total < 50] <- "<50"
  Avg_additional_charges_total.cat[Avg_additional_charges_total >= 50 & Avg_additional_charges_total < 150] <- "50-150"
  Avg_additional_charges_total.cat[Avg_additional_charges_total >= 150  & Avg_additional_charges_total < 250] <- "150-250"
  Avg_additional_charges_total.cat[Avg_additional_charges_total >= 250] <- ">250"
   } )
```

```{r}
km_fit_transplant <- survfit(surv_object ~ Avg_additional_charges_total.cat, data = gym_data)
#summary(km_fit_transplant) # commented out to reduce the pages
ggsurvplot(km_fit_transplant, conf.int = TRUE, ggtheme = theme_minimal(),
           title = "Survival Curve by Transplant Status with Censoring", 
           censor.shape = '|', censor.size=4)
```
